---
title: "DTF via Robust Scaling: Replication Code for Simulation 1"
author: Blinded
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
---

# Initialization

```{r}
library(robustDIF)
library(dplyr)
library(tidyr)
library(ggplot2)
library(see)

check_archive_robustDIF <- function(version = "0.2.0.20260708",
                                    archive.name = "review-archive") {
  installed_version <- as.character(utils::packageVersion("robustDIF"))
  if (installed_version != version) {
    stop(
      "This report expects robustDIF ", version,
      ", but found robustDIF ", installed_version,
      ". Install robustDIF from the archive branch.",
      call. = FALSE)
  }

  desc <- utils::packageDescription("robustDIF")
  installed_archive <- desc[["Config/robustDIF/archive-name"]]
  if (!identical(installed_archive, archive.name)) {
    stop(
      "This report expects robustDIF archive '", archive.name,
      "', but found archive label '", installed_archive,
      "'. Install robustDIF from the archive branch.",
      call. = FALSE)
  }
  invisible(installed_version)
}

check_archive_robustDIF()

find_results_dir <- function(candidates, pattern, min_files) {
  candidates <- unique(candidates)
  n_files <- vapply(
    candidates,
    function(path) {
      if (!dir.exists(path)) return(0L)
      length(list.files(path, pattern = pattern, full.names = TRUE))
    },
    integer(1))

  if (!any(n_files >= min_files)) {
    stop(
      "Could not find the Simulation 1 result files. Expected at least ",
      min_files, " files matching ", pattern, " in one of: ",
      paste(candidates, collapse = ", "),
      call. = FALSE)
  }

  normalizePath(candidates[which(n_files >= min_files)[1]], mustWork = TRUE)
}

sim1_pattern <- "^sim1_n[0-9]+_ndif[0-9]+[.]rds$"
results_dir <- find_results_dir(
  c(
    "../results/sim1",
    "results/sim1",
    "../../paper/results/sim1",
    "../../../paper/results/sim1",
    "paper/results/sim1"),
  pattern = sim1_pattern,
  min_files = 33L)

method_levels <- c("LRT", "RDIF", "True")

method_cols <- c(
  "LRT" = "#254A73",
  "RDIF" = "#7BAFD4",
  "True" = "grey50")

delta_method_map <- c(
  "LRT-unit" = "LRT",
  "RDIF-BH-start" = "RDIF",
  "Oracle-unit" = "True")

item_method_map <- c(
  "LRT" = "LRT",
  "RDIF-BH-start" = "RDIF")

theme_set(theme_bw(base_size = 14))
```

# Run Sims

The production run was carried out condition-by-condition. Each
condition was saved separately as an `.rds`.

```{bash, echo=TRUE, eval=FALSE}
SIM1_REPS=500 SIM1_CORES=10 SIM1_LRT_TOL=1e-3 \
  Rscript run_sim1.R
```

Simulation 1 uses 500 replications in each cell of a
`3 x 11` grid:

- sample size per group: `200`, `350`, `500`;
- number of DIF items: `0, 1, ..., 10`;
- total number of items: `20`;
- per-item DIF magnitude: `0.5` on the difficulty scale used in
  `data_sim1()`;
- reference latent mean and variance fixed at `0` and `1`;
- comparison latent mean sampled from `U(-0.25, 0.25)`;
- comparison latent variance sampled from `U(0.5, 1.5)`.

The LRT screening step uses joint slope/intercept item tests with
`TOL = 1e-3`; this was done to improve run time and piloting showed it had a neglible effect on convergence. The separate single-group `mirt(..., SE = TRUE)` fits
used by RDIF are left at the `mirt` defaults.

# Load Results

```{r}
sim_files <- list.files(
  results_dir,
  pattern = sim1_pattern,
  full.names = TRUE)

sim_list <- lapply(sim_files, readRDS)
sim_delta_all <- bind_rows(lapply(sim_list, `[[`, "delta")) |>
  mutate(
    raw_method = as.character(method),
    n_persons = n_persons_1,
    n_dif = as.integer(n_dif))

sim_delta <- sim_delta_all |>
  filter(raw_method %in% names(delta_method_map)) |>
  mutate(
    method = factor(unname(delta_method_map[raw_method]), levels = method_levels))

sim_items_all <- bind_rows(lapply(sim_list, `[[`, "items")) |>
  mutate(
    raw_method = as.character(method),
    n_persons = n_persons_1,
    n_dif = as.integer(n_dif))

sim_items <- sim_items_all |>
  filter(raw_method %in% names(item_method_map)) |>
  mutate(
    method = factor(unname(item_method_map[raw_method]),
                    levels = c("LRT", "RDIF")))

run_info <- bind_rows(lapply(sim_list, function(x) {
  data.frame(
    n.reps = x$run$n.reps,
    n.cores = x$run$n.cores,
    base.seed = x$run$base.seed,
    condition.seed = x$run$condition.seed,
    lrt.TOL = x$run$lrt.TOL,
    save.mle = x$run$save.mle,
    completed = x$run$completed)
}))

status_file <- file.path(results_dir, "sim1_status.csv")
sim_status <- if (file.exists(status_file)) read.csv(status_file) else NULL

n_reps <- length(unique(sim_delta$rep))
n_items <- unique(sim_delta$n_items)
alpha <- 0.05
```

# Item-Level Diagnostics

```{r}
item_diag <- sim_items |>
  summarize(
    flagged_rate = mean(flagged, na.rm = TRUE),
    mean_p = mean(p_value, na.rm = TRUE),
    mean_p_adjusted = mean(p_adjusted, na.rm = TRUE),
    .by = c(method, n_persons, n_dif, true_dif)) |>
  mutate(
    diagnostic = ifelse(true_dif, "True positive rate", "False positive rate"))

knitr::kable(item_diag, digits = 3)
```

```{r}
ggplot(
  item_diag,
  aes(x = n_dif, y = flagged_rate, color = method, group = method)) +
  geom_hline(yintercept = alpha, linetype = 3, color = "grey50") +
  geom_point(size = 2.2) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = 0:10) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  facet_grid(
    diagnostic ~ n_persons,
    labeller = labeller(n_persons = function(x) paste0("N = ", x))) +
  labs(
    x = "Number of Items with DIF",
    y = "Item Flagging Rate",
    color = "Item Test")
```

# Bias and Precision

## Scaling Function Bias

Figure 2 from the paper. `Naive` is the mean of all
item-level scaling functions. The other methods use the method-specific
estimate of the common-item scaling function.

```{r}
scale_bias <- bind_rows(
  sim_delta |>
    filter(method == "RDIF") |>
    transmute(
      Method = "Naive",
      n_persons,
      n_dif,
      Bias = naive_est - delta_0),
  sim_delta |>
    transmute(
      Method = as.character(method),
      n_persons,
      n_dif,
      Bias = scale_est - delta_0)) |>
  mutate(
    Method = factor(
      Method,
      levels = c("Naive", method_levels),
      ordered = TRUE))

scale_cols <- c("Naive" = "grey80", method_cols)

ggplot(scale_bias, aes(x = n_dif, y = Bias, fill = Method)) +
  geom_hline(yintercept = 0, linetype = 3, linewidth = 0.9, color = "grey40") +
  see::geom_violinhalf(
    aes(group = interaction(n_dif, Method)),
    position = position_dodge(width = 0.85),
    trim = TRUE,
    width = 2,
    linewidth = 0.25,
    alpha = 0.85) +
  stat_summary(
    aes(group = interaction(n_dif, Method)),
    fun = median,
    geom = "point",
    position = position_dodge(width = 0.85),
    size = 0.9,
    color = "black") +
  scale_x_continuous(
    breaks = 0:10,
    sec.axis = sec_axis(
      ~ . * 0.5 / 20,
      breaks = (0:10) * 0.5 / 20,
      labels = sprintf("%.2f", (0:10) * 0.5 / 20),
      name = "True Delta")) +
  scale_fill_manual(values = scale_cols) +
  facet_wrap(
    ~ n_persons,
    nrow = 3,
    scales = "free_y",
    labeller = labeller(n_persons = function(x) paste0("Sample Size = ", x))) +
  labs(
    x = "Number of Items with DIF",
    y = "Scaling Estimate - True Scaling Value",
    fill = "Method") +
  theme(panel.grid.minor = element_blank())
```

```{r}
scale_bias_table <- scale_bias |>
  summarize(
    bias = mean(Bias, na.rm = TRUE),
    .by = c(Method, n_persons, n_dif)) |>
  mutate(Method = as.character(Method)) |>
  arrange(n_persons, n_dif, Method) |>
  pivot_wider(names_from = Method, values_from = bias)

knitr::kable(scale_bias_table, digits = 3)
```

## DTF Effect Bias

```{r}
delta_bias <- sim_delta |>
  mutate(Bias = delta_hat - true_Delta) |>
  summarize(
    bias = mean(Bias, na.rm = TRUE),
    .by = c(method, n_persons, n_dif))

ggplot(delta_bias, aes(x = n_dif, y = bias, color = method, group = method)) +
  geom_hline(yintercept = 0, linetype = 3, color = "grey50") +
  geom_point(size = 2.4) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(
    breaks = 0:10,
    sec.axis = sec_axis(
      ~ . * 0.5 / 20,
      breaks = (0:10) * 0.5 / 20,
      labels = sprintf("%.2f", (0:10) * 0.5 / 20),
      name = "True Delta")) +
  scale_color_manual(values = method_cols) +
  facet_wrap(
    ~ n_persons,
    nrow = 3,
    labeller = labeller(n_persons = function(x) paste0("Sample Size = ", x))) +
  labs(
    x = "Number of Items with DIF",
    y = "DTF Estimate - True DTF",
    color = "Method")
```

```{r}
delta_bias_wide <- delta_bias |>
  mutate(method = as.character(method)) |>
  arrange(n_persons, n_dif, method) |>
  select(method, n_persons, n_dif, bias) |>
  pivot_wider(names_from = method, values_from = bias)

knitr::kable(delta_bias_wide, digits = 3)
```

# Rejection Rates

```{r}
sim_power <- sim_delta |>
  summarize(
    rejection_rate = mean(rejected, na.rm = TRUE),
    .by = c(method, n_persons, n_dif)) |>
  mutate(
    rejection_rate = ifelse(
      method == "True" & n_dif == 0,
      NA_real_,
      rejection_rate))

ggplot(sim_power, aes(x = n_dif, y = rejection_rate, color = method, group = method)) +
  geom_hline(yintercept = 0.05, linetype = 3, linewidth = 0.9, color = "grey40") +
  geom_hline(yintercept = 0.80, linetype = 3, linewidth = 0.9, color = "grey40") +
  geom_point(size = 2.4, na.rm = TRUE) +
  geom_line(linewidth = 0.8, na.rm = TRUE) +
  scale_x_continuous(
    breaks = 0:10,
    sec.axis = sec_axis(
      ~ . * 0.5 / 20,
      breaks = (0:10) * 0.5 / 20,
      labels = sprintf("%.2f", (0:10) * 0.5 / 20),
      name = "True Delta")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .25)) +
  scale_color_manual(values = method_cols) +
  facet_wrap(
    ~ n_persons,
    nrow = 3,
    labeller = labeller(n_persons = function(x) paste0("Sample Size = ", x))) +
  labs(
    x = "Number of Items with DIF",
    y = "Rejection Rate",
    color = "Method")
```

```{r}
sim_power_wide <- sim_power |>
  pivot_wider(names_from = method, values_from = rejection_rate)

knitr::kable(sim_power_wide, digits = 3)
```

# Coverage

```{r}
sim_coverage <- sim_delta |>
  summarize(
    coverage = mean(covered, na.rm = TRUE),
    mean_width = mean(2 * qnorm(.975) * delta_se, na.rm = TRUE),
    mean_se = mean(delta_se, na.rm = TRUE),
    .by = c(method, n_persons, n_dif))

ggplot(sim_coverage, aes(x = n_dif, y = coverage, color = method, group = method)) +
  geom_hline(yintercept = 0.95, linetype = 3, linewidth = 0.9, color = "grey40") +
  geom_point(size = 2.4) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(
    breaks = 0:10,
    sec.axis = sec_axis(
      ~ . * 0.5 / 20,
      breaks = (0:10) * 0.5 / 20,
      labels = sprintf("%.2f", (0:10) * 0.5 / 20),
      name = "True Delta")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .25)) +
  scale_color_manual(values = method_cols) +
  facet_wrap(
    ~ n_persons,
    nrow = 3,
    labeller = labeller(n_persons = function(x) paste0("Sample Size = ", x))) +
  labs(
    x = "Number of Items with DIF",
    y = "Coverage",
    color = "Method")
```

```{r}
sim_coverage_wide <- sim_coverage |>
  select(method, n_persons, n_dif, coverage) |>
  pivot_wider(names_from = method, values_from = coverage)

knitr::kable(sim_coverage_wide, digits = 3)
```
