---
output:
  github_document:
    toc: False
---

```{r setup, include=FALSE}
options(digits = 4)
knitr::opts_chunk$set(echo = TRUE)
```

# robustDIF

An R package that provides functions for assessing differential item functioning (DIF) in item response theory (IRT) models using methods from robust statistics. Based on the paper: Halpin, P.F. (2022) Differential Item Functioning Via Robust Scaling (https://arxiv.org/abs/2207.04598). 

Recently, `robustDIF` was updated to support the graded response model (GRM) in addition to the (2PL) IRT model, in two independent groups. It will also provide sensible results for unidimesional item factor analysis with a probit link for the measurement model and WLS estimation instead of maximum likelihood. The convenience function `get_model_parms` was updated to read-in from both `mirt` and `lavaan`. New functions were added to test whether DIF affects conclusions about impact (aka differential test functioning; `delta_test`). 

Ongoing developments focus on comparisons across more than 2 groups, dealing with dependent groups (e.g., longitudinal invariance), and omnibus tests of DIF that correspond to different "level" of measurement invariance. 

# Installation

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("peterhalpin/robustDIF")
library(robustDIF)
```

```{r, echo = F}
library(robustDIF)
```

# Example Dataset

The main user-facing functions are illustrated below using the built-in example dataset `rdif.eg`. In the example dataset, there are a total of five binary items and the first item has DIF on the intercept (threshold) equal to 1/2 SD on the latent trait. The latent trait was generated from *N*(0,1) in the reference group and *N*(.5, 1) in the comparison group.

Check out the documentation for `rdif.eg` and `get_model_parms` for more info about how to format data for use with `robustDIF`.

# The RDIF Procedure

The RDIF procedure is a robust version of "mean/mean" IRT-based scaling (see Kolen & Brennan, 2014, chap. 6). It is implemented as a post-estimation step following separate calibrations of a focal IRT model in a reference group (*g = 0*) and a comparison group (*g = 1*). The RDIF procedure uses IRT scaling parameters that are functions of the parameters of the distributions of the latent trait. Letting $\mu$ and $\sigma^2$ denote the mean and variance of the latent trait in the comparison group, the scaling parameter used to test for DIF on the item intercepts is $\theta = \mu/\sigma$ and the item-level scaling functions are $Y_i = (d_{1} - d{0})/ a_1$ where $a$ is the item slope, $d$ is the item intercept, and subscripts denote groups. The scaling parameter used to test for DIF on the item slopes is $\theta = \sigma$ and the item-level scaling functions are $Y_i = a_1/ a_0$

The `rdif` function estimates the IRT scaling parameters by applying a robust estimator of location to the item-level scaling functions. As a by-product, this estimator flags items with DIF at the desired asymptotic Type I Error rate (`alpha`). In the context of DIF, we are mainly interested in the flagging procedure. In the context of scaling, this amounts to a robust version of mean/mean scaling using the slope-intercept parameterization of an IRT model. 

In the output below, the estimated values of the scaling parameters are indicated by `est`. Items with DIF are indicated by `weights = 0`. The other output describes the iteratively re-weighted least squares estimation routine (number of iterations and the convergence criterion).

```{r}
# Item intercepts
rdif(mle = rdif.eg, par = "intercept", alpha = .05)

# Item slopes
rdif(mle = rdif.eg, par = "slope", alpha = .05)
```

We can see that the first item exhibits DIF on the intercept (i.e., it has a weight of zero). The estimated scaling parameters are reasonable approximations of the data-generating values. 

# "Stand-Alone" Wald Tests of DIF

Inferences about DIF can also be made by following up `rdif` with stand-alone Wald tests of the item parameters. The stand-alone tests can be useful if one wishes to test for DIF using a different Type I Error rate than was used for estimation with `rdif`. Otherwise, the parameter-level tests provide the same inference as the weights in the `rdif` output (albeit in a more commonly used format). 

To test each item parameter separately, use the function `rdif_z_test`:

```{r}
# Wald test of item intercepts
rdif_z_test(mle = rdif.eg, par = "intercept")

# Wald test of item slopes
rdif_z_test(mle = rdif.eg, par = "slope")
```

Alternatively, the user can test all parameters for each item together using a Wald test whose degrees of freedom is equal the number of item parameters. These item-level tests (rather parameter-level tests) are not equivalent to the flagging procedure because they combine information across parameter estimates from the same item. 

```{r}
# Wald test of both parameters
rdif_chisq_test(mle = rdif.eg)
```

In the example, the Wald tests lead to the same conclusions as the flagging procedure. 

# Does DIF Affect Impact? 

One may wish to know whether any DIF, if present, affects conclusions about impact (i.e., how the groups differ on the latent trait). This question is related to the topic of differential test functioning (DTF), which addresses the extent to which test scores are biased by DIF (e.g., Chalmers et al., 2016). One way to  approach this question is to compare a naive estimate of impact that ignores DIF to an estimator that is robust to DIF. In `robustDIF`, the naive estimator is chose to be the MLE of the scaling parameter, computed from the item-level scaling functions, which is a precision-weighted version of the standard "mean-mean" approach. The robust estimator is as above (see Halpin, 2022, for details). The null hypothesis that both estimators are consistent for the "true" scaling parameter leads to a Hausman-like specification test of DTF. 

The `delta_test` function implement this test. Note that, unlike other procedures for DTF, implementation of this test does not require an initial item-by-item analysis to identify which items may exhibit DIF -- it allows the user to test for DTF *before* doing a DIF analysis. 


```{r}
# Comparing Naive and Robust estimates to evaluate DTF on the means of the latent trait
delta_test(mle = rdif.eg, par = "intercept")

# For the variance of latent trait
delta_test(mle = rdif.eg, par = "slope")
```

In the example, it can be seen that large DIF on a single item's intercept did not lead to statistically different conclusions about impact. 

# The Rho function

For data analyses, it is useful to check whether the M-estimator of the IRT scaling parameters has a clear global minimum before proceeding to make inferences about DIF. In `robustDIF`, the minimization problem is described by `rho_fun`.

```{r}
# Rho function for item intercepts
rho.intercept <- rho_fun(mle = rdif.eg, par = "intercept", grid.width = .01)
par(mfrow = c(1,2))
plot(rho.intercept$theta, rho.intercept$rho, type = "l")

# Rho function for item slopes
rho.slope <- rho_fun(mle = rdif.eg, par = "slope", grid.width = .01)
plot(rho.slope$theta, rho.slope$rho, type = "l")
```

Note the minimizing values of theta in the plots corresponds to the values reported by `rdif` above.

# References

Chalmers, R. P., Counsell, A., & Flora, D. B. (2016). It might not make a big DIF: Improved differential test functioning statistics that account for sampling variability. Educational and Psychological Measurement, 76(1), 114–140. https://doi.org/10.1177/0013164415584576

Halpin, P.F. (2022) Differential Item Functioning Via Robust Scaling. https://arxiv.org/abs/2207.04598. 

Kolen, M. J., & Brennan, R. L. (2014). Test Equating, Scaling, and Linking. Springer. https://doi.org/10.1007/978-1-4939-0317-7

